Flaring star-planet interactions in Kepler and TESS

### 2023-06-15 src/tex/ms.tex
[showyourwork] overleaf sync

HIP 67522 is a young Sun, currently contracting onto the main sequence. It is a Sco-Cen member (10-20 Myr old), which was discovered to host a close-in Jupiter, HIP 67522 b, in 2020~\citep{rizzuto2020tess}. Curiously, the planet is in close spin-orbit commensurability -- $P_{\rm rot}/P_{\rm orb}\approx 1/5$. Therefore, we cannot completely rule out that the observed periodicity is in fact a rotational periodicity, and not related to the planet. However, if we calculate the $p$-value of the AD test using the rotational period instead of the orbital period, we find HIP 67522 deviates from uniform flaring in time with $p=0.18$. So we cannot rule rotational modulation completely, since the rotational period is very uncertain relative to the total observing baseline, and the clustering close to transit is suggestive.% Fortunately, HIP 67522 will be observed again with TESS in April 2023, which may resolve the puzzle.

@@

\subsubsection{KOI-12}
\label{sec:results:individualstars:koi12}
KOI-12, also known as Kepler-448, is a two-planet system with an eccentric outer planet KOI-12 c, and an inner Warm Jupiter planet KOI-12 b in a $17.8\,$d orbit~\citep{masuda2017eccentric}. With respect to rotation rate, stellar and planetary radius, the KOI-12 system is similar to HIP~67522 and TAP~26 (although for the those young stars the radii are large because the stars are still contracting onto the main sequence). In constrast, KOI-12 is $1.4\pm0.3\,$Gyr old~\citep{bourrier2015sophie}, an F5 subgiant~\citep{frasca2016activity}, which is already evolving off the main sequence. Yet, despite its high expected power of interaction, we detect no excess flaring in phase with KOI-12 b. Here, the reason could be that KOI-12 b is in fact super-Alfv\'enic due to the relatively wide orbit compared to TAP 26 and HIP 67522. However, our estimate of the Alf\'en radius place KOI-12 b in the sub-Alfv\'enic regime~(see Fig.~\ref{fig:as}), although the scaling law applied there may be inaccurate for post-main sequence stars. Intermittent interaction might play a role here as well, and could be favored if future observations capture episodes during which the interaction is 'on'.

@@

In \cite{reiners2022magnetism} it was reported that for Sun-like stars with a total magnetic field strength of up to 2000~G, the longitudinal magnetic field derived from Stokes V spectropolarimetric observations ($B_{\rm V}$) accounts for $\sim$\,$10$\% of the total magnetic field. \cite{kochukhov2020hidden} showed that this trend also holds for M-dwarfs with magnetic fields up to 4000~G. In this case, however, the recovered longitudinal magnetic field accounts for about 20\% of the total magnetic field. Moreover, our star sample covers a wide range of magnetic fields, the weakest being 100~G and the strongest 3800~G. Therefore, as a first assumption, we consider that the longitudinal magnetic field of our stars would reach 15\% of the total magnetic field (i.e. $B_{\rm V} = 0.15B$), given that most systems in our sample have B~$>$~2000 G.

@@

We show the difference between star-planet distance and $\overline{AS}_{R}$ in Fig.~\ref{fig:as}. HIP 67522 b falls most deeply into the sub-Alfv\'enic regime, consistent with the strongest signal of magnetic SPI in our sample. TAP 26 and KOI-12 also appear sub-Alfv\'enic, despite absent signal of magnetic SPI. From this follows that the absence of interaction is not caused by the planets orbiting outside the Alfv\'en radius. location inside $\overline{AS}_{R}$ is a required but not sufficient condition for magnetic SPI, as we will see in the following Sections. %\textcolor{red}{JDAG: Thats a good point. However, I think being inside the AS is a required but not sufficient condition for the type of SPI that you are looking for. Because you are searching for flares, you need reconnection + particle acceleration (which are the precursors/drivers of the flare emission observed on different wavelengths). While the coronal material is expected to provide the particles (at least in closed-field Active Regions -- if the reconnection takes place in open-field line regions, the plasma density there is usually low which would also yield weak EM signatures as those are typically a function of $\rho$), one of the difficulties I see is that in the star-planet system, the reconnection of the magnetic field is more likely to occur in the planet end rather than on the stellar end (simply because the AR field strength is probably larger than the planetary field). As such, the field topology will be more 'resistant' closer to the star than to the planet, facilitating the re-organization in the latter case compared to the former. Planetary outflows falling back to the star could help to trigger some instabilities, but those would depend on whether the planet is losing any significant mass (which we probably don't know for these targets). There's of course more to it than just field strength or accretion, with things like the torsion and helicity of the field among others (which is why the likelihood of a solar AR to host a flare is not only determined by field strength but also by its topology/complexity), but all of that is too complicated or even impossible to estimate here. The bottom line is that there are multiple reasons that could explain why those two stars that lie deep in the sub-Alfv\'enic regime show no signs of flaring SPI which, without a better characterization of each individual system + some 3D modelling, are very hard to ascertain.}

@@

In our results in Fig.~\ref{fig:adtest_bp}, we can interpret the two branches emerging at high expected power of interaction as due to intermittency. TAP 26 and KOI-12 do not show any sign of excess flaring in our data, despite similar expected powers as in HIP 67522 with its $2-3\sigma$ signal. However, other explanations cannot be ruled out yet (see Sections~\ref{sec:results:individualstars:tap26} and \ref{sec:results:individualstars:koi12}, and the following).

@@

\paragraph{HIP 67522} appears with tentative signs of both tidal and magnetic interaction (measured and expected in the different scenarios). On the one hand, since the flares in HIP 67522 mostly appear close to the transit phase, and not during eclipse, this may favor magnetic interaction, and the tidal signal would only be an artifact. On the other hand, the narrow range in which most of the flares on HIP 67522 occurred~(see Fig.~\ref{fig:cumdist_transiting} and Section~\ref{sec:results:individualstars:hip67522}) might favor a well-localized active region in the vicinity of the tidal bulge. Conceivably, both effects play a role at the same time.

@@

We also searched the same data for signs of locally expressed tidal interaction, that is, excess flaring in phase with the tidal bulges ($P_{\rm orb}/2$). We found a similar trend with different scaling laws of tidal interaction, albeit at even lower significance. Our system-by-system comparison suggests that young systems, like HIP 67522, may interact both tidally and magnetically, whereas old, magnetically inactive systems, like Kepler-42, might mostly interact tidally.

Our study is the first that systematically searched for flaring star-planet interactions, and is not limited by sampling or selection bias in the way earlier studies of this effect have often been. While our results remain tentative, future studies using our technique will benefit from accumulating data. The TESS mission continues to produce high precision light curves for many nearby active planet hosts, and the PLATO~\citep{rauer2014plato} mission will join in late 2026. For AU Mic, we may be only a few TESS Sectors away from a $3\sigma$ level confirmation~\citep{ilin2022searching} of the tentative interaction signal observed here and in other work~\citep{klein2022one}. Several systems in our sample are already scheduled for further observing with TESS, including HIP 67522, Kepler-42, and KOI-12.

>>>>

HIP 67522 is a young Sun, currently contracting onto the main sequence. It is a Sco-Cen member (10-20 Myr old), which was discovered to host a close-in Jupiter, HIP 67522 b, in 2020~\citep{rizzuto2020tess}. Curiously, the planet is in close spin-orbit commensurability -- $P_{\rm rot}/P_{\rm orb}\approx 1/5$. Therefore, we cannot completely rule out that the observed periodicity is in fact a rotational periodicity, and not related to the planet. The $p$-value of the AD test using the rotational period instead of the orbital period is $p=0.18$. However, the rotational period is very uncertain relative to the total observing baseline (coherence time of about 100 days, and observing baseline of several years), so this number is unreliable.

@@

\subsubsection{KOI-12}
\label{sec:results:individualstars:koi12}
KOI-12, also known as Kepler-448, is a two-planet system with an eccentric outer planet KOI-12 c, and an inner Warm Jupiter planet KOI-12 b in a $17.8\,$d orbit~\citep{masuda2017eccentric}. With respect to rotation rate, stellar and planetary radius, the KOI-12 system is similar to HIP~67522 and TAP~26 (although for the those young stars the radii are large because the stars are still contracting onto the main sequence). In constrast, KOI-12 is $1.4\pm0.3\,$Gyr old~\citep{bourrier2015sophie}, an F5 subgiant~\citep{frasca2016activity}, which is already evolving off the main sequence. Yet, despite its high expected power of interaction, we detect no excess flaring in phase with KOI-12 b. The reason could be that KOI-12 b is in fact super-Alfv\'enic due to the relatively wide orbit compared to TAP 26 and HIP 67522. However, our estimate of the Alf\'en radius place KOI-12 b in the sub-Alfv\'enic regime~(see Fig.~\ref{fig:as}), although the scaling law applied there may be inaccurate for post-main sequence stars. Intermittent interaction might play a role here as well, and could be favored if future observations capture episodes during which the interaction is 'on'.

@@

In \cite{reiners2022magnetism} it was reported that for Sun-like stars with a total magnetic field strength of up to 2000~G, the longitudinal magnetic field derived from Stokes V spectropolarimetric observations ($B_{\rm V}$) accounts for $\sim$\,$10$\% of the total magnetic field. \cite{kochukhov2020hidden} showed that this trend also holds for M-dwarfs with magnetic fields up to 4000~G. In this case, however, the recovered longitudinal magnetic field accounts for about $20\%$ of the total magnetic field. Moreover, our star sample covers a wide range of magnetic fields, the weakest being 100~G and the strongest $3800\,$G. Therefore, as a first assumption, we consider that the longitudinal magnetic field of our stars would reach 15\% of the total magnetic field (i.e. $B_{\rm V} = 0.15B$), given that most systems in our sample have $B>2000\,$G.

@@

We show the difference between star-planet distance and $\overline{AS}_{R}$ in Fig.~\ref{fig:as}. HIP 67522 b falls most deeply into the sub-Alfv\'enic regime, consistent with the strongest signal of magnetic SPI in our sample. K2-25, TOI-540 and AU Mic also fulfil the condition of sub-Alfv\'enic orbit of the innermost planet, consistent with their observed marginal deviation from uniform flaring with orbital phase.

TAP 26 and KOI-12 also appear sub-Alfv\'enic, despite absent signal of magnetic SPI. From this follows that the absence of interaction is not caused by the planets orbiting outside the Alfv\'en radius. Location inside $\overline{AS}_{R}$ is a required but not sufficient condition for magnetic SPI, as we will see in the following Sections. %\textcolor{red}{JDAG: Thats a good point. However, I think being inside the AS is a required but not sufficient condition for the type of SPI that you are looking for. Because you are searching for flares, you need reconnection + particle acceleration (which are the precursors/drivers of the flare emission observed on different wavelengths). While the coronal material is expected to provide the particles (at least in closed-field Active Regions -- if the reconnection takes place in open-field line regions, the plasma density there is usually low which would also yield weak EM signatures as those are typically a function of $\rho$), one of the difficulties I see is that in the star-planet system, the reconnection of the magnetic field is more likely to occur in the planet end rather than on the stellar end (simply because the AR field strength is probably larger than the planetary field). As such, the field topology will be more 'resistant' closer to the star than to the planet, facilitating the re-organization in the latter case compared to the former. Planetary outflows falling back to the star could help to trigger some instabilities, but those would depend on whether the planet is losing any significant mass (which we probably don't know for these targets). There's of course more to it than just field strength or accretion, with things like the torsion and helicity of the field among others (which is why the likelihood of a solar AR to host a flare is not only determined by field strength but also by its topology/complexity), but all of that is too complicated or even impossible to estimate here. The bottom line is that there are multiple reasons that could explain why those two stars that lie deep in the sub-Alfv\'enic regime show no signs of flaring SPI which, without a better characterization of each individual system + some 3D modelling, are very hard to ascertain.}

@@

In our results in Fig.~\ref{fig:adtest_bp}, we can interpret the two branches emerging at high expected power of interaction as due to intermittency. TAP 26 and KOI-12 do not show any sign of excess flaring in our data, despite similar expected powers as in HIP 67522 with its $>2.5\sigma$ signal. However, other explanations cannot be ruled out yet (see Sections~\ref{sec:results:individualstars:tap26} and \ref{sec:results:individualstars:koi12}, and the following).

@@

\paragraph{HIP 67522} does not show signs of tidal interaction, despite a relatively high expected power. Conceivably, both effects play a role at the same time, but the tidal interaction is dominated by magnetic SPI. With more available observations in the future, and better constraints on the expected signal of each type of SPI, it will become possible to model both simultaneously. However, with only a total of 12 flares in this work, we cannot make any strong claims about the contribution of tidal SPI, except for it apparently being weaker than the magnetic aspect.

@@

We also searched the same data for signs of locally expressed tidal interaction, that is, excess flaring in phase with the tidal bulges ($P_{\rm orb}/2$). We found a similar trend with different scaling laws of tidal interaction, albeit at even lower significance. Our system-by-system comparison suggests that young systems, like HIP 67522, may be dominated by magnetic interaction, whereas old, magnetically inactive systems, like Kepler-42, might mostly interact tidally.

Our study is the first that systematically searched for flaring star-planet interactions, and is not limited by sampling or selection bias in the way earlier studies of this effect have often been. While our results remain tentative, future studies using our technique will benefit from accumulating data. The TESS mission continues to produce high precision light curves for many nearby active planet hosts, and the PLATO~\citep{rauer2014plato} mission will join in late 2026. For AU Mic, we may be only a few TESS Sectors away from a $3\sigma$ level confirmation~\citep{ilin2022searching} of the tentative interaction signal observed here and in other work~\citep{klein2022one}. Several systems in our sample are already scheduled for further observing with TESS, including Kepler-42, and KOI-12.
